---
title: 'REPRODUCIBLE RESEARCH : WEEK 2 ASSIGNMENT'
author: "Niharika Gupta"
date: "26/04/2020"
output: html_document
---


##Loading and Preprocessing the Data

1.  Load the data 

To load the data,I use the read.csv command:
```{r}
activity<-read.csv("activity.csv")
```

To know how the data looks like , I use str(),summary() and head() commands:
```{r}
str(activity)
```

```{r}
summary(activity)
```

```{r}
head(activity)
```

2. Process/transform the data (if necessary) into a format suitable for your analysis

For the following task , we will need to remove the missing values :
```{r}
act.complete<-na.omit(activity)
```

##What is the mean total number of steps taken per day?

1.  Calculate the total number of steps taken per day

I use tapply function to calculate the total number of steps per day 

```{r}
act.day<-tapply(act.complete$steps,act.complete$date,sum)
```

There should be one obseravation per day now:

```{r}
summary(act.day)
```

2.  Make a histogram of the total number of steps taken each day

```{r}
hist(act.day,xlab="Total number of steps",col="black")
```

3.  Calculate and report the mean and median of the total number of steps taken per day

I use the mean() and median() functions:

```{r}
mean(act.day["steps"])
```

```{r}
median(act.day["steps"])
```

##What is the average daily activity pattern?

1.  Make a time series plot (i.e. type = "1")of the 5-minute interval(x-axis) and the average number of steps taken, averaged across all days(y-axis) 

I use the dplyr library to collpase the data by day, creating the sum of steps:

```{r}
library(dplyr)
act.int<-group_by(act.complete,interval)
act.int<-summarize(act.int,steps=mean(steps))
```

Next we plot the average daily steps against the intervals:

```{r}
library(ggplot2)
ggplot(act.int,aes(interval,steps)) + geom_line()
```

2.  Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?

We find the row in the interval data frame for which steps is equal to the maximum number of steps, then we look at the interval of that row:

```{r}
act.int[act.int$steps==max(act.int$steps),]
```

##Imputing Missing Values

Note that there are a number of days/intervals where there are missing values (coded as NA). The presence of missing days may introduce bias into some calculations or summaries of the data.

1.  Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)

The total number of rows with NAs is equal to the difference between the number of rows in the raw data and the number of rows in the data with only complete cases:

```{r}
nrow(activity)-nrow(act.complete)
```

2.  Devise a strategy for filling in all of the missing values in the dataset. The strategy does not need to be sophisticated. For example, you could use the mean/median for that day, or the mean for that 5-minute interval, etc.

I replace missing values with the mean number of steps for each interval across all of the days. The act.int data frame contains these means. I start by merging the act.int data with the raw data:

```{r}
names(act.int)[2] <- "mean.steps"
act.impute <- merge(activity, act.int)
```

3.  Create a new dataset that is equal to the original dataset but with the missing data filled in.

I replace the value with the mean number of steps for the interval across days:

```{r}
act.impute$steps[is.na(act.impute$steps)] <- act.impute$mean.steps[is.na(act.impute$steps)]
```

4.  Make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day. Do these values differ from the estimates from the first part of the assignment? What is the impact of imputing missing data on the estimates of the total daily number of steps?

I first create a dataset with the total number of steps per day using the imputed data:

```{r}
act.day.imp <- group_by(act.impute, date)
act.day.imp <- summarize(act.day.imp, steps=sum(steps))
```

Then I generate the histogram and summary statistics:

```{r}
qplot(steps , data=act.day.imp , col = "black")
```

```{r}
mean(act.day.imp$steps)
```

```{r}
median(act.day.imp$steps)
```
The mean appears to be unaffected by this simple data imputation. The median is smaller.

##Are there differences in activity patterns between weekdays and weekends?

For this part the weekdays() function may be of some help here. Use the dataset with the filled-in missing values for this part.

Create a new factor variable in the dataset with two levels – “weekday” and “weekend” indicating whether a given date is a weekday or weekend day.

I convert the date variable to the date class, then use the weekdays() function to generate the day of the week of each date. I create a binary factor to indicate the two weekend days:

```{r}
act.impute$dayofweek <- weekdays(as.Date(act.impute$date))
act.impute$weekend <-as.factor(act.impute$dayofweek=="Saturday"|act.impute$dayofweek=="Sunday")
levels(act.impute$weekend) <- c("Weekday", "Weekend")
```

2.  Make a panel plot containing a time series plot (i.e. type = “l”) of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis).

First I create separate data frames for weekends and weekdays:

```{r}
act.weekday <- act.impute[act.impute$weekend=="Weekday",]
act.weekend <- act.impute[act.impute$weekend=="Weekend",]
```

Then for each one, I find the mean number of steps across days for each 5 minute interval:

```{r}
act.int.weekday <- group_by(act.weekday, interval)
act.int.weekday <- summarize(act.int.weekday, steps=mean(steps))
act.int.weekday$weekend <- "Weekday"
act.int.weekend <- group_by(act.weekend, interval)
act.int.weekend <- summarize(act.int.weekend, steps=mean(steps))
act.int.weekend$weekend <- "Weekend"
```

I append the two data frames together, and I make the two time series plots:

```{r}
act.int <- rbind(act.int.weekday, act.int.weekend)
act.int$weekend <- as.factor(act.int$weekend)
ggplot(act.int, aes(interval, steps)) + geom_line() + facet_grid(weekend ~ .)
```






